% ===== AAAI-27 spine draft: Related Work (~0.9 col, expanded) =====
\section{Related Work}\label{sec:related}

\paragraph{Logit and tuned lenses.} The logit
lens~\citep{nostalgebraist2020logitlens} projects intermediate residual
streams through a language model's frozen unembedding; the tuned
lens~\citep{belrose2023tuned} fits a small per-layer affine map instead.
Both were built for autoregressive text models, where the unembedding
gives every layer a shared, well-defined target vocabulary; a DiT
editor has no such vocabulary; a patch's hidden state is not a
distribution over anything until the whole network has run. We port
both lens families to this setting -- staying on the frozen-head side
for honesty and crossing to the learned side only to render the
internal image -- with the linear representation
hypothesis~\citep{park2024linear} licensing the informative
disagreement between a frozen head and a probe.

\paragraph{Instruction-based diffusion image editing.} Two lines
established how a diffusion model can be made to follow an edit
instruction at all. InstructPix2Pix~\citep{brooks2023instructpix2pix}
trains directly on synthesized instruction-image pairs so the model
learns the edit end to end; Prompt-to-Prompt~\citep{hertz2023prompttoprompt}
instead manipulates the model's cross-attention maps at inference time
to inject the edit while preserving the source's structure, training-free.
Both contribute a \emph{mechanism for producing} the edit; neither asks
what representation the network forms once an instruction is given, or
where in the forward pass it becomes the pixels that change. Our lens
targets that open question on exactly such a production editor -- not a
new editing mechanism, but a reading of the one it already has.

\paragraph{Thinking with images, and diffusion interpretability.} A
growing line has multimodal models reason \emph{with} explicit
intermediate images by externalizing a visual scratchpad
\citep{hu2024sketchpad,wu2024vot,li2025mvot}. We supply the mechanistic
complement: an image that exists \emph{latently} inside a generative editor before any
output pixel is produced, with no explicit scratchpad or tool call involved.
The Diffusion Lens~\citep{toker2024diffusionlens} decodes a
text-to-image model's \emph{text encoder}; DiffLens and
ICM~\citep{shi2025difflens,zaleska2026icm} probe and steer internal
features, including choices absent from an underspecified prompt; DreamReader
unifies extraction, patching, steering, and stitching for text-to-image
U-Nets~\citep{sykora2026dreamreader}; and sparse autoencoders expose causal
generation features~\citep{surkov2025sae}. Breaking the Lock-in finds early
cross-seed convergence in text-to-image sampling and intervenes to increase
diversity~\citep{zhu2026lockin}. Probe-Select~\citep{guo2026probeselect}
predicts final quality from early activations;
ELECT~\citep{kim2025elect} scores early latent background consistency;
and ADE-CoT~\citep{qu2026adecot} adds adaptive budgets
and stopping at scale. We instead predict \emph{which outcome} a fixed
underspecified edit realizes and use selection only as a held-out behavioral
test, while mapping how an image-editing DiT's realized outcome changes code
across depth; we do not infer irreversible lock-in. We decode an
editing DiT's own denoising
stack as an image across both layer and time, building on the flow-matching
and rectified-flow formulation~\citep{lipman2023flowmatching,liu2023rectifiedflow}
and classifier-free guidance~\citep{ho2022cfg} that
\model{}~\citep{qwen2025image,peebles2023dit} uses.

\paragraph{Causal interpretability and editing.} Causal tracing / ROME~\citep{meng2022rome}
patches activations between runs, the move behind our transplant;
\citet{heimersheim2024activation} catalogue patching pitfalls that
motivate our cautious interpretation of ablation. Recent analyses further show
that detector-friendly diffusion features can produce artifacts when steered
off distribution~\citep{pegoraro2026look}, and that patching with multiple
mediators cannot by itself isolate a unique causal pathway
\citep{zhang2026curse}. Accordingly, transplant is used only as a sufficiency
test and steering as a bounded write test. Finally,
activation addition~\citep{turner2023actadd} is the steering primitive we
instantiate with a probe direction. All three were built for language
models with a token vocabulary at every patched site; we port them to a
continuous, spatial image stream, where sufficiency must be checked
against a rendered image rather than a next-token distribution. Concept
Lancet~\citep{luo2025colan} already uses compositional representation
transplant for diffusion editing; our transplant is therefore an
interventional sufficiency check within a layer--time map, not a standalone
editing-method claim. Our
semantic readout uses CLIP~\citep{radford2021clip} for lack of a better
zero-shot alternative.
