% ===== AAAI-27 spine draft: Method (~1.3 col-page target) =====
\section{Proposed Method: A Layer--Time Lens and Intervention Probes}\label{sec:method}

Figure~\ref{fig:method} summarizes the full pipeline; here we specify each
component. We study \model{}, a 60-block DiT editor whose image stream
concatenates noisy target tokens with always-clean condition tokens. All
quantities below are read from the
\emph{post-block residual} of selected blocks via forward hooks,
captured in the model's native \texttt{bfloat16}: image-stream
activations reach absmax ${\sim}3{\times}10^{9}$ even at layer 0, so a cast
to \texttt{float16} silently overflows most elements to \texttt{inf}
(bf16 capture is a lossless copy, not a downcast).

\paragraph{Notation and layer--time readout.}
Let $h_{\ell,t}^{(i)}\!\in\!\mathbb{R}^{d}$ denote the post-block residual
after block $\ell\!\in\!\{0,\ldots,59\}$ at denoising step $t$, for
image-stream row $i$ (target patch, condition patch, or a hooked subset);
$h^{\mathrm{tar}}_{\ell,t}$ stacks the target-patch rows used for all
image-space readouts below (condition rows are captured only for explicitly
stated controls). A lens operator
$\mathcal{L}_{\ell,t}:\mathbb{R}^{n\times d}\!\to\!\mathbb{R}^{n\times64}$
maps these residuals to a target-grid velocity estimate $\tilde v_{\ell,t}$.
Logit- and Jacobian-style lenses in language models align internal states
with output units along depth and token position; a DiT patch has no
vocabulary, so we define a \emph{layer--time lens} as a map from
$(\ell,t)$ to an image and then a score on a fixed grid:
\begin{equation}
  I_{\ell,t}
  = \mathcal{V}\!\left(\Pi_t\!\left(
  \mathcal{L}_{\ell,t}(h^{\mathrm{tar}}_{\ell,t})\right)\right),
  \qquad
  \Pi_t(\tilde v)=x_t-\sigma_t\tilde v,
  \label{eq:lens-grid}
\end{equation}
where $\mathcal{V}$ is the frozen VAE decoder. We score a fixed edit-region
crop of $I_{\ell,t}$ with CLIP ViT-L/14~\citep{radford2021clip}. With
unit-normalized image and prompt-ensembled text embeddings $e_I,e_T$, and
CLIP's learned logit scale $\gamma$, the forced-choice probability over the
reported label bank $\{y_k\}$ is
\begin{equation}
  [\mathcal{M}_{\ell,t}]_k
  \;=\;
  \mathrm{softmax}_k\!\Bigl(
    \gamma\, e_T(y_k)^\top e_I(\mathrm{crop}(I_{\ell,t}))
  \Bigr).
  \label{eq:clip-readout}
\end{equation}
Figures~\ref{fig:teaser} and~\ref{fig:method} instantiate
\eqref{eq:lens-grid}--\eqref{eq:clip-readout}: the image-editor analogue of
lexical alignment grids in logit/Jacobian-lens visualizations, not a full
$\partial \mathbf{y}/\partial h$ (a donor-scalar Jacobian basis is tested in
the supplement and does not coincide with probe or write directions).

\paragraph{Lens operators.}
The \emph{time lens} uses the terminal head output at step $t$:
\begin{equation}
  \hat{x}_0(x_t,t) = x_t - \sigma_t\, v_\theta(x_t,t),
  \label{eq:time-lens}
\end{equation}
decoded through the VAE---what the model currently believes it will output.
The \emph{depth lens} asks the same question at intermediate depth by
re-invoking the model's frozen output head (non-affine
\texttt{AdaLayerNormContinuous} + linear \texttt{proj\_out}, bit-exact
whether invoked at layer 59 or earlier):
\begin{equation}
  v_{\ell,t} = \mathrm{Head}(h^{\mathrm{tar}}_{\ell,t}),
  \qquad
  \hat{x}_{0,\ell,t} = x_t - \sigma_t\, v_{\ell,t},
  \label{eq:depth-lens}
\end{equation}
then $I_{\ell,t}=\mathcal{V}(\hat{x}_{0,\ell,t})$.
Because \texttt{norm\_out} has no learned scale, rescaling $h^{\mathrm{tar}}_{\ell,t}$ to
the terminal norm is a no-op (raw vs.\ rescaled agree to
\texttt{max\_rel\_diff}$=0$); we use the raw head everywhere with a
standing self-consistency gate on every grid.
The \emph{tuned lens}~\citep{belrose2023tuned} fits a per-$(\ell,t)$ ridge map
from $d{=}3072$ hidden dimensions to the $64$-d conditional-pass velocity,
one dictionary per cell,
\begin{equation}
  \hat{v}_{\ell,t} = W_{\ell,t} h^{\mathrm{tar}}_{\ell,t} + b_{\ell,t},
  \quad
  W_{\ell,t} = V^\top H (H^\top H + \lambda I)^{-1},
  \label{eq:tuned-lens}
\end{equation}
after token-wise centering; $\lambda$ is chosen per cell on a held-out
$10\%$ \emph{token} split, distinct from zero-shot tests that hold out an
entire edit family and source image (hyperparameters in the supplement).
The \emph{differential lens} differences edit and control runs that share a
seed (step-0 latents are then bit-identical):
\begin{equation}
  D_{\ell,t}
  = \bigl|I^{\mathrm{edit}}_{\ell,t}
  - I^{\mathrm{ctrl}}_{\ell,t}\bigr|,
  \label{eq:diff-lens}
\end{equation}
isolating the internal edit plan when the unchanged prompt is a faithful
no-op; a control-fidelity gate on correlation with the source flags cases
where it is not.

\paragraph{Linear probes.}
Neither \eqref{eq:clip-readout} nor a lens image is itself a causal claim.
We additionally fit a lightweight probe
$p(y \mid h_{\ell,t}) = \mathrm{softmax}(W h_{\ell,t})$ on hidden states
(cross-seed evaluation), reading the same $(\ell,t)$ grid in label space
rather than pixel space---the basis mismatch against the frozen head in
\eqref{eq:depth-lens} is central to our mechanism results.

\paragraph{Intervention probes.}
Three forward-hook overwrites of the post-block residual test distinct
properties, applied identically to both CFG passes:
\begin{align}
  h &\leftarrow \mathbf{0}\ \text{or}\ \bar{h}
  && \text{(ablate: necessity stress test)}, \nonumber \\
  h^{(\mathrm{rec})} &\leftarrow h^{(\mathrm{don})}
  && \text{(transplant: sufficiency)}, \nonumber \\
  h &\leftarrow h + \alpha d
  && \text{(steer: linear writability)},
  \label{eq:interventions}
\end{align}
with $\alpha$ in units of per-row RMS and $d$ a probe direction; an empty
ablation spec reproduces the baseline bit-for-bit. Repeated ablation can
drive activations off-manifold, so we treat it only as a coarse stress test.

\paragraph{Protocol: predict, look, gate.} Three rules make the
\emph{reasoning} reproducible, not just the runs. \textbf{Predict, then
run}: every intervention was pre-registered; refutations are reported as
prominently as confirmations, and several headline results are
pre-registered predictions that failed informatively. \textbf{Look at
every image}: no verdict rests on CLIP alone -- every run carries a
per-image visual verdict recorded before the aggregates were read, and
where the two disagree the image wins. \textbf{Gate the instrument}: the
empty-ablation and raw-vs-rescaled checks are standing regressions re-run
per experiment. Porting the method to another editor needs only a
native-dtype residual hook, the frozen-head gate, and region-restricted
probes -- the recipe executed unchanged in the cross-model replication
below.
