% ===== AAAI-27 spine draft: Translated late / two codes (~1.6 col) =====
\section{The Edit Is Translated Late: Two Codes}\label{sec:translate}

If the decision is present by layer 6 but the frozen head cannot read it
until layer 52, the mid-stack must hold the edit in a \emph{different
code}. Three measurements show what that code is and where it is
converted.

\paragraph{Two decoding conventions cross over.} At an intermediate layer
the frozen head yields a per-token vector $v_\ell$ we normally treat as a
velocity, forming $\hat x_0 = x_t - \sigma_t v_\ell$ (variant A). Decoding
$v_\ell$ instead \emph{directly}, as if it were already a clean latent
(variant C), inverts the story: for \emph{car\_red} at step 4, at L24 the
standard decode A scores $P(\text{red}){=}0.12$ (still wrong-colored)
while the direct decode C already scores $0.66$; at L59 this reverses (A
$=0.93$, C $=0.02$). The mid-stack is not blank -- it encodes the target
content in an \emph{image-like} convention, and a narrow band near the top
reconciles it to the velocity convention the head consumes. (This is a
whole-vector reweighting, not a literal sign flip: essentially no
individual channel of $v_\ell$ has a negative slope against $v_{59}$, mean
$0.003$.)

\paragraph{The conversion is a sharp, timestep-invariant band.} A dense
every-layer sweep from L44 to L59 (Fig.~\ref{fig:boundary}) locates the
A/C crossover precisely: transition midpoint L53.6 at step 16 (width one
layer) and L52.7 at step 4, the two differing by under one layer -- the
boundary's \emph{location} is timestep-invariant. Crucially, the
cross-seed probe accuracy is close to \emph{flat} ($0.77$--$0.84$) across
every layer from L44 through L58 and drops only at the terminal layer L59
($0.68$--$0.70$). So the translation that changes what the head can read
(L52--54) and the loss of linear separability (L59) are two \emph{distinct}
events sharing a neighborhood: the target-image code survives the
translation intact and is degraded only by the final, velocity-specific
projection.

\paragraph{A tuned lens renders the internal image, layer by layer.} To
confirm the mid-stack code is a genuine \emph{image} rather than a probe
artifact, we fit a per-(layer, timestep) tuned lens
(\S\ref{sec:method}) on \emph{remove}/\emph{background}/\emph{addition}
edits and evaluate it \emph{zero-shot} on the held-out \emph{car\_red}
color family. Held-out $R^2$ rises with depth from $0.73$--$0.89$ at L6 to
$0.998$ at L59, where a standing identity gate confirms the target is
correct (decoding L59 through the frozen head reproduces the true velocity
bit-for-bit, max$|$diff$|{=}0$). Every depth down to L6 decodes a coherent
scene: a pre-registered prediction that readability itself would have a
boundary is refuted -- the L52--54 boundary is a property of the frozen
head's one fixed projection, not of the representation. At step 0 the
pixels are pure noise, so the decode comes entirely from the velocity
code, and we watch the edit enter the internal image directly
(Fig.~\ref{fig:tunedlens}): the source's blue car persists through L24,
turns orange-red by L36, the scene composes by L44, and the car is fully
red by L52. The same frozen dictionaries transfer with zero retraining to
six further held-out cases (style, more colors, removal, background,
addition); \emph{add\_runner} even reproduces the delayed onset on the
internal image's own time axis -- an empty path at step 0, a blurred
figure by step 2, a clear runner by step 4.

\paragraph{The anti-image, mechanized.} This also explains a striking
artifact: the mid-stack \emph{frozen}-head decode at later steps is a
color-\emph{inverted} ghost -- a cyan car under a red-brown sky, the exact
complement of the true scene. It is the image-like mid-stack code forced
through a head built to consume a velocity: $\hat x_0 = x_t - \sigma v$
flips the sign of an image-like quantity. The tuned lens at the same cell
is already coherent.
