% ===== AAAI-27 spine draft: Differential lens + 15% preview (~0.85 col-page) =====
\section{A Differential Lens: Previewing the Edit at 15\% NFE}\label{sec:difflens}

The tuned lens renders the whole internal image; to isolate the
\emph{edit} from the reconstruction we difference it against a control.
We run the same source and seed twice -- once with the edit prompt, once
with the literal instruction ``keep the image unchanged'' -- and decode
both through the \emph{same} frozen dictionaries (no refit). Since the two
runs share a seed, their step-0 latents are bit-identical, so
$D_{\ell,t}=|\text{decode}(h_\ell^{\text{edit}})-\text{decode}(h_\ell^{\text{unchanged}})|$
isolates the internal edit plan, the reconstruction cancelling by
construction. This also answers \S\ref{sec:translate}'s open caveat that
the tuned lens's high $R^2$ might be easy background variance: token-level
$R^2$ \emph{inside} the edit region is $0.88$--$0.97$, matching the
full-image value -- the dictionaries carry the edit-specific signal
itself.

\paragraph{Commitment is early; separation is gradual.} A pre-registered
prediction that $D$ would already localize tightly at every step is
refuted, informatively. The edit \emph{content} is present at step 0 (the
tuned lens already shows red eyes, a black trunk), but the
\emph{differential} at step 0 is a diffuse whole-frame field -- the edit
prompt perturbs the entire velocity relative to the no-op, not just the
edited region -- and condenses onto the edit's footprint only gradually
(localization of $D$'s top mass at L52 climbs monotonically across steps,
e.g.\ $0.21{\to}0.47{\to}0.67{\to}0.83$ for a cat's-eyes recolor).
\emph{Committed} (is the content in the internal image) and
\emph{separated} (has the differential found where it lives) are two
different quantities, and the value of the no-op control is that it also
caught a case where the model \emph{cannot} do nothing -- one flat-art
scene rendered a transparent-PNG checkerboard for ``keep unchanged,''
which we discard, motivating a control-fidelity gate before any
differential number is trusted.

\paragraph{A preview lever.} Because the differential carries the edit's
footprint, we ask whether it does so early enough to be useful. At step 2
-- the third of 20, $\approx$15\% of the forward passes -- across a
10-case, five-family battery of natural photographs (all passing the
control-fidelity gate, $r\in[0.86,0.99]$), the median IoU between
$D_{\text{L52},t2}$'s top mass and the true edit region is $0.33$,
$6.6\times$ chance, and all 10 cases sharpen from step 0 to step 2
(Fig.~\ref{fig:previewlever}): an object slated for removal lights up as a
ghost, a keep-vs-replace background segmentation emerges unsupervised, and
a tiny recolor glows at its exact site. (Honestly, a fixed top-10\%
ground-truth mask under-scores genuinely tiny edits, padding their region
${\sim}10\times$ with collateral divergence -- a sixth documented
metric-vs-eye divergence in this work, and the first not caused by CLIP.)
Because the no-op control is \emph{independent of the edit prompt}, it can
be computed once per source image and cached; previewing where and how
large a \emph{candidate} edit will land then costs only the three forward
passes to reach step 2 plus one frozen-dictionary decode, not a full
20-step run -- a training-free way to triage many candidate prompts
against one image before committing to any.
