% ===== AAAI-27 spine draft: Generalize + lever (~0.9 col) =====
\section{Does It Generalize? A Cross-Model Check}\label{sec:generalize}

Every finding so far concerns one checkpoint. We re-ran the depth-lens
crossover, the cross-seed probe, and layer-band ablation on two further
checkpoints sharing \model{}'s transformer classes: FireRed-Image-Edit-1.1,
an \emph{independent fine-tune}, and Rapid-AIO, a \emph{4-step distilled}
merge. \textbf{The translation boundary and the early decision replicate.}
FireRed's A/C crossover lands at L53--56, Rapid-AIO's at L52--54 -- both on
\model{}'s L52--54 band; FireRed's probe reads $0.852$ at L6 and dips to
$0.69$ at L59, the same flat-then-dip shape. \textbf{The decision direction
itself transfers across models}: a probe trained on one model and tested on
the \emph{other} reaches $0.84$/$0.92$ accuracy, and the two models'
L12 probe directions have cosine similarity $0.833$ -- the edit lives in
nearly the same direction of the shared 3072-d space, not merely the same
layer index. The full tuned-lens toolchain also ports to FireRed (deep-layer
$R^2$ $0.97$--$0.999$, anti-image reproduced). \textbf{One honest
divergence}, caught by the check doing its job: ablating L48--59 leaves
\model{}'s edit intact ($0.826$) but is catastrophic for FireRed ($0.035$)
-- late-stack redundancy is a property of \emph{training}, not the
architecture, so \S\ref{sec:dissoc}'s readable$\neq$necessary holds
architecture-generally only in its early half.

\section{A Lever: Truncating Classifier-Free Guidance}\label{sec:lever}

Because the decision is linearly present by layer 6 and translated by
layer 53, there is no architectural reason classifier-free guidance needs
both of its forward passes at all 20 steps. Running the unconditional pass
only for the first two steps and the conditional pass alone thereafter
leaves all four edit families visually intact at a \textbf{45\% reduction
in forward passes} (a 20-case battery confirms it, mean $1.77\times$
speedup). Dropping CFG entirely isolates its real job: recolor, removal,
and addition still land from the conditional pass alone, but background
swap composites the new scene \emph{around} un-erased source structure
($P(\text{beach}){=}0.45$ vs.\ ${\ge}0.996$) -- CFG's non-redundant
contribution is resolving \emph{scope}, not committing the edit. The same
decision structure survives $7.7\times$ step distillation (20/20 cases
land), so distillation compresses the sampling trajectory but not the
decision structure the lens exposes.
