% ===== AAAI-27 spine draft: Limits + Conclusion (~0.8 col) =====
\section{Limitations and Honest Interpretation}\label{sec:limits}

This tool is exploratory. Every number comes primarily from one model,
one or two seeds, and a double-digit count of hand-chosen images; the
cross-model checks (\S\ref{sec:generalize}) cover two checkpoints of the
same lineage on a subset of measurements, and we make no claim about
``intent'' or about recovering the network's ``true'' reasoning. The
semantic readout inherits CLIP's failure modes in identifiable ways: pure
sampling noise can be mis-scored (a still-noisy decode labeled beach at
$0.795$), and, in mirror image, an obviously-red recolored car
\emph{under}-scores ($0.09$--$0.16$) while a \emph{failed} flat patch
scores higher ($0.70$--$0.84$) -- which is why every quantitative claim is
paired with a pre-registered visual verdict, and the image wins ties. Four
of our own initial misreadings are retained in the text as data: a
thumbnail ``revision'' phenomenon that quantification erased (leaving a
cleaner printing-vs-thinking split), a tempting continuous novelty law
that did not survive (only a categorical one did), a steering ``clean
flip'' that was a flat patch replacing the car's geometry, and the
over-strong conclusion we drew from it -- that a direction cannot carry
shape -- which a silhouette mask refuted. We claim only that, on \model{},
the readable/necessary/writable dissociations we report are real and
repeatable for the seeds and prompts used, because we checked them
against causal interventions rather than passive readouts alone.

\section{Conclusion}\label{sec:conclusion}

We built a layer--time and tuned lens for a 60-block DiT image editor and
paired every reading with a causal check. The edit is decided early
(linearly readable by layer 6), carried as a target-image code for most of
the depth, and translated into the velocity code only in a sharp,
timestep-invariant band at layer $\approx$53 -- a boundary that survives an
independent fine-tune and a 4-step distillation, while readability,
necessity, and writability dissociate three distinct ways. If one sentence
organizes the paper, it is this: \emph{the lens shows the translation, not
the decision -- reading the decision requires changing basis, and verifying
it requires intervention.} Behind it stands the picture we opened with and
have now earned clause by clause: an edit is not developed gradually -- it
is decided early, bound to a spatial carrier, and translated late into the
strokes that paint it. A generative editor natively thinks with an internal
image before it makes one; the lens turns that thought into something that
can be watched, steered, and budgeted for.
